MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-22 docs/paper/out/paper.tex
add @marina-manrique corrections, some of my own

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Alexey Alekhin {et~al.}} %use et al only if is more than 1 author
\def\Authors{
 Alexey Alekhin\(^\dagger\)\,$^{1}$  Evdokim Kovach\(^\dagger\)\,$^{1}$  Marina Manrique\,$^{1}$  Pablo Pareja\,$^{1}$  Eduardo Pareja\,$^{1}$  Raquel Tobes\,$^{1}$  and Eduardo Pareja-Tobes $^{1,*}$
}% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Oh no sequences! Research Group, Era7 Bioinformatics, Granada, Spain}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Corresponding Author}
\def\corrAddress{Oh no sequences! Research Group, Era7 Bioinformatics, Plaza Campo Verde 3, Granada, 18001, Spain}
\def\corrEmail{eparejatobes@ohnosequences.com}

@@

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and/or frequently substituted
by new ones or by databases specifically designed for focused projects.
MG7 uses parallelization and distributed analysis based on AWS (Amazon
Web Services), with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations designed for AWS that
counts with a static reproducible specification of dependencies and
behavior of the different components. The modeling of the taxonomy tree
is done using the new paradigm of graph databases of Bio4j that
facilitates the taxonomic assignment tasks and the calculation of the
taxa abundance values considering the hierarchic structure of taxonomy
tree. MG7 includes the new 16S-DB7 database built with a flexible and
sustainable system of updating and project-driven personalization.

@@

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using \emph{Statika} and \emph{Datasets}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using \emph{Loquat}, a composable
  system for scaling/parallelizing stateless computations especially
  designed for Amazon Web Services (AWS)
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j). It facilitates the taxonomic assignment tasks and
  the calculation of the taxa abundance values considering the
  hierarchic structure of taxonomy tree (cumulative values)
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms Lowest Common Ancestor and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

The input and output (and their locations) being defined statically has
several critical advantages. First, composing different loquats is easy
and safe; just use the output types and locations of the first one as
input for the second one. Second, data and their types help in not
mixing different resources when implementing a process, while serving as
a safe and convenient mechanism for writing generic processing tasks.
For example, merging paired-end Illumina reads generically is easy as
the data type includes the relevant information (insert size, read
length, etc) to pass to a tool such as FLASH.

\subsubsection{Type-safe eDSLs for BLAST and
FLASH}\label{type-safe-edsls-for-blast-and-flash}

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASH} and
\href{https://github.com/ohnosequences/blast}{BLAST} expressions and
their execution.

@@

In the same spirit as for BLAST, we implemented a type-safe eDSL for
FLASH expressions and their execution, supporting features equivalent to
those outlined for the BLAST eDSL.

@@

Our 16S Reference Database is a curated subset of sequences from NCBI
nucleotide database \textbf{nt}. The sequences included were selected by
similarity with the bacterial and archaeal reference sequences
downloaded from the \textbf{RDP database} \citep{cole2013ribosomal}. RDP
unaligned sequences were used to capture new 16S sequences from
\textbf{nt} using BLAST similarity search strategies and then,
performing additional curation steps to remove sequences with poor
taxonomic assignments to taxonomic nodes close to the root of the
taxonomy tree. All the nucleotide sequences included in \textbf{nt}
database has a taxonomic assignment provided by the \textbf{Genbank}
sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.

@@

\subsubsection{Joining reads of each pair using
FLASH}\label{joining-reads-of-each-pair-using-flash}

In the first step the paired-end reads, designed with an insert size
that yields pairs of reads with an overlapping region between them, are
assembled using FLASH \citep{magovc2011flash}. FLASH is designed to
merge pairs of reads when the original DNA fragments are shorter than
twice the length of reads. Thus, the sequence obtained after joining the
2 reads of each pair is larger and has better quality since the sequence
at the ends of the reads is refined merging both ends in the assembly.
To have a larger and improved sequence is crucial to do more precise the
inference of the bacterial origin based on similarity with reference
sequences.

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity (by default \(evalue \leq e^{-15}\))
filtering those hits that are not sufficiently good comparing them with
the best one. We select the best HSP (High Similarity Pair) per
reference sequence and then choose the best HSP (that with lowest e
value) between all the selected ones. The bitscore of this best HSP
(called S) is used as reference to filter the rest of HSPs. All the HSPs
with bitscore below p x S are filtered. p is a coefficient fixed by the
user to define the bitscore required, e.g.~if p=0.9 and S=700 the
required bitscore threshold would be 630. Once we have the definitive
HSPs selected, we obtain their corresponding taxonomic nodes using the
taxonomic assignments that NCBI provides for all the nt database
sequences. Now we have to analyze the topological distribution of these
nodes in the taxonomy tree: i. If all the nodes forms a line in the
taxonomy tree (are located in a not branched lineage to the tree root)
we should choose the most specific taxID as the final assignment for
that read. We call to this kind of assignment the `in line' exception
(see Figure 2 right panel). ii. If not, we should search for the
\emph{sensu stricto} Lowest Common Ancestor (LCA) of all the selected
taxonomic nodes (See Figure 2 left panel). In this approach we decided
to use the bitscore for evaluating the similarity because it is a value
that increases when similarity is higher and depends a lot on the length
of the HSP. Some reads could not find sequences with enough similarity
in the database and then they would be classified as reads with no hits.
Advanced metagenomics analysis approaches \citep{huson2013microbial}
have adopted LCA assignment algorithms because it provides fine and
trusted taxonomical assignment.

@@

In order to be able to exploit Amazon Web Services infrastructure for
the MG7 analysis, the user needs to set up an AWS account with certain
IAM (Identity and Access Management) permission policies that will grant
access to the resources used in the workflow.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Treat data analysis as a software project. This makes for radical
  improvements in \emph{reproducibility}, \emph{reuse},
  \emph{versioning}, \emph{safety}, \emph{automation} and
  \emph{expressiveness}
\item
  input and output data, their locations and type are expressible and
  checked at compile-time using \emph{Datasets} ­3. management of
  dependencies and machine configurations using \emph{Statika}
\item
  automation of AWS cloud resources and processes, including
  distribution and parallelization through the use of \emph{Loquat}
\item
  taxonomic data and related operations are treated natively as what
  they are: graphs, through the use of \emph{Bio4j}
\item
  MG7 provides a sustainable model for taxonomic assignment, appropriate
  to face the challenging amount of data that high throughput sequencing
  technologies generate
\end{enumerate}

@@

\begin{itemize}
\tightlist
\item
  \textbf{Reproducibility} the same analysis can be run again with
  exactly the same configuration in a trivial way.
\item
  \textbf{Versioning} as in any software project, there can be different
  versions, stable releases, etc.
\item
  \textbf{Reuse} we can build standard configurations on top of this and
  reuse them for subsequent data analysis. A particular data analysis
  \emph{task} can be used as a \emph{library} in further analysis.
\item
  \textbf{Decoupling} We can start working on the analysis
  specification, without any need for available data in a much easier
  way.
\item
  \textbf{Documentation} We can take advantage of all the effort put
  into software documentation tools and practices, such as in our case
  Scaladoc or literate programming. As documentation, analysis processes
  and data specification live together in the files, it is much easier
  to keep coherence between them.
\item
  \textbf{Expresiveness and safety} For example in our case we can
  choose only from valid Illumina read types, and then build a default
  FLASH command based on that. The output locations, being declared
  statically, are also available for use in further analysis.
\end{itemize}

@@

An important aspect of the MG7 workflow is the way it deals with data
resources. All the data that is going to be used in the analysis or
produced as an output is described as Scala code using rich types from
the \emph{Datasets} language. This allows user to specify all the
information about the type of the data that can be utilized then by the
tools analyzing this data. For example, we can specify that for the
first part of the MG7 workflow running FLASH in parallel, requires
Illumina paired end reads and produces joined reads.

On one hand, specification of the input data allows us to restrict its
type and force users to be conscious about what they pass as an input.
On the other hand specification of the output data helps to build a
workflow as a \emph{composition} of several parts: we can ensure on the
Scala code type level that the output of one component fits as an input
for the next component. This is crucial, as obviously the way a data
analysis task works depends a lot on the particular structure of the
data. For instance, in the MG7 workflow, using BLAST eDSL, we can
precisely describe which format will have the output of the BLAST step,
which information it will include, and then in the next step we can
reuse this description to parse BLAST output and retrieve the part of
the information needed for the taxonomy assignment analysis. Having data
structure described statically as Scala code allows us to be sure that
we won't have parsing problems or other issues with incompatible data
passed between workflow components.

@@

Every external tool involved in the workflow is represented as a
\emph{Statika} bundle, which is essentially a Scala project describing
the installation process of this tool and declaring dependencies on
other bundles which will be installed prior to the considered tool
itself. Describing relationships between bundles on the code level
allows us to track the directed acyclic graph of their dependencies and
linearize them to automatically install them sequentially in the right
order. Meanwhile describing installation process on the code level
allows user to utilize wide range of Scala and Java APIs, making
installation a well-defined sequence of steps rather than an unreliable
script dependent on the certain environment. This way \emph{Statika}
provides an easy way to make deployment an automated reproducible
process.

Besides the bioinformatics tools like BLAST and FLASH, \emph{Statika}
bundles are used for wrapping data dependencies and all inner components
of the system that require cloud deployment. In particular, all
components of \emph{Loquat} are bundles; the user can then define which
components are needed for the parallel processing on each computation
unit in an expressive way, declaring them as bundle dependencies of the
loquat ``worker'' bundle. This modularization is also important for the
matter of making components of the system reusable for different
projects and liberating the user from most of the tasks related to their
deployment.

@@

The MG7 workflow consists of certain steps, each of which perform some
work in parallel, using the cloud infrastructure managed by
\emph{Loquat}. It is important to notice the horizontal scalability of
this approach. Irrespectively of how much data needs to be processed,
MG7 will easily handle it, by splitting data on chunks and performing
the analysis on multiple computation units. The Amazon Elastic Compute
Cloud (EC2) service provides a transparent way of managing computation
infrastructure, called autoscaling groups. The User can set MG7
configuration parameters, adjusting for each task the amount and
hardware characteristics of the EC2 instances they want to occupy for
it. But it's important to note that as each workflow step is not very
resource demanding, it is not needed to hire EC2 instances with some
advanced hardware, instead an average type will work and you can win
time by simply scaling the number of the hired instances.

@@

It is certainly possible to adapt MG7 to work with shotgun metagenomics
data. Simply changing the reference database to include whole genome
sequence data could yield interesting results. This could also be
refined by restricting reference sequences according to all sort of
criteria, like biological function or taxonomy. Bio4j would be an
invaluable tool here, thanks to its ability to express express complex
predicates on sequences using all the information linked with them (GO
annotations, UniProt data, NCBI taxonomy, etc).

@@

\subsection{FLASH eDSL}\label{flash-edsl}

MG7 uses
\href{https://github.com/ohnosequences/flash/releases/tag/v0.1.0}{ohnosequences/flash
v0.1.0}. The FLASH version used is v1.2.11.

>>>>

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Alexey Alekhin {et~al.}} %use et al only if is more than 1 author
\def\Authors{
 Alexey Alekhin\(^\dagger\)\,$^{1}$  Evdokim Kovach\(^\dagger\)\,$^{1}$  Marina Manrique\,$^{1}$  Pablo Pareja-Tobes\,$^{1}$  Eduardo Pareja\,$^{1}$  Raquel Tobes\,$^{1}$  and Eduardo Pareja-Tobes $^{1,*}$
}% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Oh no sequences! Research Group, Era7 Bioinformatics, Granada, Spain}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Eduardo Pareja-Tobes}
\def\corrAddress{Oh no sequences! Research Group, Era7 Bioinformatics, Plaza Campo Verde 3, Granada, 18001, Spain}
\def\corrEmail{eparejatobes@ohnosequences.com}

@@

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and/or frequently substituted
by new ones or by databases specifically designed for focused projects.
MG7 uses parallelization and distributed analysis based on Amazon Web
Services (AWS), with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations designed for AWS that
counts with a static reproducible specification of dependencies and
behavior of the different components. The modeling of the taxonomy tree
is done using the new paradigm of graph databases of Bio4j that
facilitates the taxonomic assignment tasks and the calculation of the
taxa abundance values considering the hierarchic structure of taxonomy
tree. MG7 includes the new 16S-DB7 database built with a flexible and
sustainable system of updating and project-driven personalization.

@@

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using \emph{Statika} and \emph{Datasets}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using \emph{Loquat}, a composable
  system for scaling/parallelizing stateless computations especially
  designed for AWS
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j). It facilitates the taxonomic assignment tasks and
  the calculation of the taxa abundance values considering the
  hierarchic structure of taxonomy tree (cumulative values)
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms Lowest Common Ancestor and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

The input and output (and their locations) being defined statically has
several critical advantages. First, composing different loquats is easy
and safe; just use the output types and locations of the first one as
input for the second one. Second, data and their types help in not
mixing different resources when implementing a process, while serving as
a safe and convenient mechanism for writing generic processing tasks.
For example, merging paired-end Illumina reads generically is easy as
the data type includes the relevant information (insert size, read
length, etc) to pass to a tool such as FLASh.

\subsubsection{Type-safe eDSLs for BLAST and
FLASh}\label{type-safe-edsls-for-blast-and-flash}

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASh} and
\href{https://github.com/ohnosequences/blast}{BLAST} expressions and
their execution.

@@

In the same spirit as for BLAST, we implemented a type-safe eDSL for
FLASh expressions and their execution, supporting features equivalent to
those outlined for the BLAST eDSL.

@@

Our 16S Reference Database is a curated subset of sequences from NCBI
nucleotide database \textbf{nt}. The sequences included were selected by
similarity with the bacterial and archaeal reference sequences
downloaded from the \textbf{RDP database} \citep{cole2013ribosomal}. RDP
unaligned sequences were used to capture new 16S RNA sequences from
\textbf{nt} using BLAST similarity search strategies and then,
performing additional curation steps to remove sequences with poor
taxonomic assignments to taxonomic nodes close to the root of the
taxonomy tree. All the nucleotide sequences included in \textbf{nt}
database has a taxonomic assignment provided by the \textbf{Genbank}
sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.

@@

\subsubsection{Joining reads of each pair using
FLASh}\label{joining-reads-of-each-pair-using-flash}

In the first step the paired-end reads, designed with an insert size
that yields pairs of reads with an overlapping region between them, are
assembled using FLASh \citep{magovc2011flash}. FLASh is designed to
merge pairs of reads when the original DNA fragments are shorter than
twice the length of reads. Thus, the sequence obtained after joining the
2 reads of each pair is larger and has better quality since the sequence
at the ends of the reads is refined merging both ends in the assembly.
To have a larger and improved sequence is crucial to do more precise the
inference of the bacterial origin based on similarity with reference
sequences.

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a BLAST expect value threshold (by default \(evalue \leq e^{-15}\))
filtering those hits that are not sufficiently good comparing them with
the best one. We select the best HSP (High Similarity Pair) per
reference sequence and then choose the best HSP (that with lowest
E-value) between all the selected ones. The bitscore of this best HSP
(called S) is used as reference to filter the rest of HSPs. All the HSPs
with bitscore below the product \(p S\) are filtered. p is a coefficient
fixed by the user to define the bitscore required, e.g.~if \(p=0.9\) and
\(S=700\) the required bitscore threshold would be \(630\). Once we have
the definitive HSPs selected, we obtain their corresponding taxonomic
nodes using the taxonomic assignments that NCBI provides for all the nt
database sequences. Now we have to analyze the topological distribution
of these nodes in the taxonomy tree: i. If all the nodes forms a line in
the taxonomy tree (are located in a not branched lineage to the tree
root) we should choose the most specific taxID as the final assignment
for that read. We call to this kind of assignment the `in line'
exception (see Figure 2 right panel). ii. If not, we should search for
the \emph{sensu stricto} Lowest Common Ancestor (LCA) of all the
selected taxonomic nodes (See Figure 2 left panel). In this approach we
decided to use the bitscore for evaluating the similarity because it is
a value that increases when similarity is higher and depends a lot on
the length of the HSP. Some reads could not find sequences with enough
similarity in the database and then they would be classified as reads
with no hits. Advanced metagenomics analysis approaches
\citep{huson2012microbial} have adopted LCA assignment algorithms
because it provides fine and trusted taxonomical assignment.

@@

In order to be able to exploit AWS infrastructure for the MG7 analysis,
the user needs to set up an AWS account with certain IAM (Identity and
Access Management) permission policies that will grant access to the
resources used in the workflow.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Treat data analysis as a software project. This makes for radical
  improvements in \emph{reproducibility}, \emph{reuse},
  \emph{versioning}, \emph{safety}, \emph{automation} and
  \emph{expressiveness}
\item
  input and output data, their locations and type are expressible and
  checked at compile-time using \emph{Datasets}
\item
  management of dependencies and machine configurations using
  \emph{Statika}
\item
  automation of AWS cloud resources and processes, including
  distribution and parallelization through the use of \emph{Loquat}
\item
  taxonomic data and related operations are treated natively as what
  they are: graphs, through the use of \emph{Bio4j}
\item
  MG7 provides a sustainable model for taxonomic assignment, appropriate
  to face the challenging amount of data that high throughput sequencing
  technologies generate
\end{enumerate}

@@

\begin{itemize}
\tightlist
\item
  \textbf{Reproducibility} the same analysis can be run again with
  exactly the same configuration in a trivial way.
\item
  \textbf{Versioning} as in any software project, there can be different
  versions, stable releases, etc.
\item
  \textbf{Reuse} we can build standard configurations on top of this and
  reuse them for subsequent data analysis. A particular data analysis
  \emph{task} can be used as a \emph{library} in further analysis.
\item
  \textbf{Decoupling} We can start working on the analysis
  specification, without any need for available data in a much easier
  way.
\item
  \textbf{Documentation} We can take advantage of all the effort put
  into software documentation tools and practices, such as in our case
  Scaladoc or literate programming. As documentation, analysis processes
  and data specification live together in the files, it is much easier
  to keep coherence between them.
\item
  \textbf{Expresiveness and safety} For example in our case we can
  choose only from valid Illumina read types, and then build a default
  FLASh command based on that. The output locations, being declared
  statically, are also available for use in further analysis.
\end{itemize}

@@

An important aspect of the MG7 workflow is the way it deals with data
resources. All the data that is going to be used in the analysis or
produced as an output is described as Scala code using rich types from
the \emph{Datasets} language. This allows The user can specify
information about types of data, information that can then be utilized
by tools analyzing this data. For example, we can specify that for the
first part of the MG7 workflow running FLASh in parallel, requires
Illumina paired end reads and produces joined reads.

On one hand, specification of the input data allows us to restrict its
type and force users to be conscious about what they pass as an input.
On the other hand specification of the output data helps to build a
workflow as a \emph{composition} of several parts: we can ensure on the
Scala code type level that the output of one component fits as an input
for the next component. This is crucial, as obviously the way a data
analysis task works depends a lot on the particular structure of the
data. For instance, in the MG7 workflow, using BLAST eDSL, we can
precisely describe which format will have the output of the BLAST step,
which information it will include, and then in the next step we can
reuse this description to parse BLAST output and retrieve the part of
the information needed for the taxonomy assignment analysis. Having data
structure described statically as Scala code allows us to be sure that
we will not have parsing problems or other issues with incompatible data
passed between workflow components.

@@

Every external tool involved in the workflow is represented as a
\emph{Statika} bundle, which is essentially a Scala project describing
the installation process of this tool and declaring dependencies on
other bundles which will be installed prior to the considered tool
itself. Describing relationships between bundles on the code level
allows us to track the directed acyclic graph of their dependencies and
linearize them to automatically install them sequentially in the right
order. Meanwhile describing the installation process on the code level
allows the user to utilize the wide range of available Scala and Java
APIs and tools, making installation a well-defined sequence of steps
rather than an unreliable script, dependent on a certain environment.
\emph{Statika} offers an easy path towards making deployment an
automated, reproducible process.

Besides bioinformatics tools like BLAST and FLASh, \emph{Statika}
bundles are used for wrapping data dependencies and all inner components
of the system that require cloud deployment. In particular, all
components of \emph{Loquat} are bundles; the user can then define which
components are needed for the parallel processing on each computation
unit in an expressive way, declaring them as bundle dependencies of the
loquat ``worker'' bundle. This modularization is also important for the
matter of making components of the system reusable for different
projects and liberating the user from most of the tasks related to their
deployment.

@@

The MG7 workflow consists of certain steps, each of which performs some
work in parallel, using the cloud infrastructure managed by
\emph{Loquat}. It is important to notice the horizontal scalability of
this approach. Irrespectively of how much data needs to be processed,
MG7 will handle it, by splitting data into chunks and performing the
analysis on multiple computation units. The Amazon Elastic Compute Cloud
(EC2) service provides a transparent way of managing computation
infrastructure, called autoscaling groups. The User can set MG7
configuration parameters, adjusting for each task the amount and
hardware characteristics of the EC2 instances they want to use for it.
But it is important to note that, as each workflow step is not very
resource demanding, it is not needed to hire EC2 instances with some
advanced hardware, instead an average type will work and you can reduce
execution time by simply scaling out the number of instances.

@@

It is certainly possible to adapt MG7 to work with shotgun metagenomics
data. Simply changing the reference database to include whole genome
sequence data could yield interesting results. This could also be
refined by restricting reference sequences according to all sort of
criteria, like biological function or taxonomy. Bio4j would be an
invaluable tool here, thanks to its ability to express complex
predicates on sequences using all the information linked with them (GO
annotations, UniProt data, NCBI taxonomy, etc).

@@

\subsection{FLASh eDSL}\label{flash-edsl}

MG7 uses
\href{https://github.com/ohnosequences/flash/releases/tag/v0.1.0}{ohnosequences/flash
v0.1.0}. The FLASh version used is v1.2.11.
